EuPPollNet: A European database of plant-pollinator networks

### 2024-10-02 Manuscript/EuPPollNet_Ms.qmd
Minor edits on refs

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we only evaluated the taxonomic coverage of groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups (e.g., beetles), it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of pollinators across the European continent.

@@

The estimated sampling coverage of plant and pollinator species within EuPPollNet across the different networks is approximately `r round(poll_sampling_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). When we consider the accumulated pollinator richness across sampled plant species, this curve also shows a “quasi-asymptotic” growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). We find that a small portion of plant species and pollinator species are shared across a broad range of networks and that most plant (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinator (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) species are exclusively found in less than 1% of networks (**Figures 4e-4f**). The most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`* when excluding *Apis mellifera*) species are found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of networks, respectively.

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and, as expected, were not independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S4**). Although latitude has a statistically significant impact on residual connectance, it explains only a small portion of the observed variability in residual connectance and nestedness across networks. (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ < `r round(connectance_lat_pval, 2)`, **Figure 6c**; NODFc: R$^2$ ~ `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6d**). Overall, networks towards higher latitudes showed lower residual connectance but similar nestedness compared to networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate significant negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ < `r round(resid_connectance_nestedness_corr_pval, 2)`). Empirical networks did not show statistically different nestedness to the simulated ones (**Figure 6b** and **Figure S5**). The 'curveball' method for binary networks resulted in `r round(under_z_percent_curveball, 2)`% of networks statistically less nested than null expectations, `r round(no_diff_z_percent_curveball, 2)`% showing no difference, and `r round(over_z_percent_curveball, 2)`% being more nested. The 'quasiswap_count' algorithm for quantitative networks resulted in `r round(under_z_percent_quasiswap, 2)`% of networks being less nested, `r round(no_diff_z_percent_quasiswap, 2)`% showing no difference and `r round(over_z_percent_quasiswap, 2)`% being more nested. Note that in both cases, NODFc is calculated on binarised matrices.

@@

Consistent with @olesen2002, we found that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the number of possible connections that can be established between plants and pollinators for a given network size, resulting in more connected at lower latitudes in Europe. However, our results cannot be extrapolated to lower latitudes outside Europe, as tropical systems might behave differently. Finally, a result that can be surprising is that most empirical networks (~85%) showed a non-nested structure. This result contrasts with earlier work suggesting that plant-pollinator networks are typically nested [@bascompte2003] and it aligns with recent evidence showing that nestedness is not a common feature in plant-pollinator networks when evaluated against restrictive null models that conserve the observed species degree [@payrato2019\; Figure S6]. Note that while species degree distributions are sufficient to explain the emergence of nestedness, this does not preclude nestedness from being a useful metric for comparison across networks. For example, plant-pollinator networks are more nested than plant-herbivore networks precisely because these two network types differ in their composition of generalist and specialist species [@thebault2010].

@@

>>>>

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we only evaluated the taxonomic coverage of groups with species checklists available in Europe (i.e., bees, syrphids and butterflies). While there is not a good understanding of pollinator diversity in other taxonomic groups (e.g., beetles), it was assumed that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`). Therefore, the total number of of flower visiting species from other taxonomic groups (i.e., non-bee, non syrphid and non-butterfly flower visitors) at European level was extrapolated by assuming that their coverage is equal to the mean coverage of bees, syrphids, and butterflies. Consequently, we provide an estimate for the total number of potential pollinators across the European continent.

@@

The estimated sampling coverage of plant and pollinator species within EuPPollNet across the different networks is approximately `r round(poll_sampling_coverage)`% for both taxonomic groups. This suggests that the rarefied accumulation curves of both plant and pollinator species exhibit already a "quasi-asymptotic" growth of species richness by considering the current number of networks (**Figure 4a-4b**). The predicted observed species richness by doubling the sampling effort on the already sampled habitat types within the database will only increase pollinator richness by `r format(round(poll_increase3, 2), nsmall=2)`% and plant richness by `r format(round(plant_increase3, 2), nsmall=2)`%. However, the sampling coverage of interactions is `r format(round(int_coverage, 2), nsmall=2)`%, and by doubling the sampling effort, the predicted number of unique interactions recorded will have approximately a twofold increase (`r format(round(int_increase3, 2), nsmall=2)`%; **Figure 4c**). When we consider the accumulated pollinator richness across sampled plant species, this curve also shows a “quasi-asymptotic” growth with a sampling coverage value of `r format(round(plant_poll_coverage, 2), nsmall=2)`%. The predicted recorded pollinator species by doubling the number of plants sampled is expected to increase by `r format(round(plant_poll_increase3, 2), nsmall=2)`% (**Figure 4d**). We find that a small portion of plant species and pollinator species are shared across a broad range of networks and that most plant (`r format(round(plants_under_one_percent, 2), nsmall=2)`%) and pollinator (`r format(round(polls_under_one_percent, 2), nsmall=2)`%) species are exclusively found in less than 1% of networks (**Figures 4e-4f**). The most common plant (*`r most_common_plant`*) and pollinator (*`r most_common_poll`* when excluding *Apis mellifera*) species are found in `r format(round(most_common_plant_percent, 2), nsmall=2)`% and `r format(round(most_common_poll_percent, 2), nsmall=2)`% of networks, respectively.

@@

Connectance values ranged between `r round(min_connectance, 2)` to `r round(max_connectance, 2)` ($\bar{x}$ = `r round(mean_connectance, 2)`) and followed a negative exponential relationship with the number of species per network (Kendall $\tau$ = `r round(connectance_spp_mean_corr_tau, 2)`, _P_ < 0.01; **Figure 6a**). Nestedness values (NODFc) ranged between `r round(min_nodfc, 2)` to `r round(max_nodfc, 2)` ($\bar{x}$ = `r round(mean_nodfc, 2)`), and, as expected, were not independent of the mean number of species (Kendall $\tau$ = `r round(normalised_nestedness_spp_mean_corr_tau, 2)`, _P_ = `r round(normalised_nestedness_spp_mean_corr_pval, 2)`; **Figure S4**). Although latitude has a statistically significant impact on residual connectance, it explains only a small portion of the observed variability in residual connectance and nestedness across networks. (connectance: R$^2$ = `r round(connectance_lat_r2, 2)`, _P_ < `r round(connectance_lat_pval, 2)`, **Figure 6c**; NODFc: R$^2$ ~ `r round(normalised_nestedness_lat_r2, 2)`, _P_ = `r round(normalised_nestedness_lat_pval, 2)`, **Figure 6d**). Overall, networks towards higher latitudes showed lower residual connectance but similar nestedness compared to networks located at lower latitudes. Note that residual connectance and normalised nestedness showed a moderate significant negative correlation (Kendall $\tau$ = `r round(resid_connectance_nestedness_corr_tau, 2)`, _P_ < `r round(resid_connectance_nestedness_corr_pval, 2)`). Empirical networks did not show statistically different nestedness (NODF) to the simulated ones (**Figure 6b** and **Figure S5**). The 'curveball' method for binary networks resulted in `r round(under_z_percent_curveball, 2)`% of networks statistically less nested than null expectations, `r round(no_diff_z_percent_curveball, 2)`% showing no difference, and `r round(over_z_percent_curveball, 2)`% being more nested. The 'quasiswap_count' algorithm for quantitative networks resulted in `r round(under_z_percent_quasiswap, 2)`% of networks being less nested, `r round(no_diff_z_percent_quasiswap, 2)`% showing no difference and `r round(over_z_percent_quasiswap, 2)`% being more nested. Note that in both cases, NODF is calculated on binarised matrices.

@@

Consistent with @olesen2002, we found that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the number of possible connections that can be established between plants and pollinators for a given network size, resulting in more connected networks at lower latitudes in Europe. However, our results cannot be extrapolated to lower latitudes outside Europe, as tropical systems might behave differently. Finally, a result that can be surprising is that most empirical networks (~85%) showed a non-nested structure. This result contrasts with earlier work suggesting that plant-pollinator networks are typically nested [@bascompte2003] and it aligns with recent evidence showing that nestedness is not a common feature in plant-pollinator networks when evaluated against restrictive null models that conserve the observed species degree [@payrato2019\; Figure S6]. Note that while species degree distributions are sufficient to explain the emergence of nestedness, this does not preclude nestedness from being a useful metric for comparison across networks. For example, plant-pollinator networks are more nested than plant-herbivore networks precisely because these two network types differ in their composition of generalist and specialist species [@thebault2010].

@@

- Figure S6
